
```{r }
---
title: "Assignment 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}

#install.packages("caret")
#install.packages("randomForest")
#install.packages("rpart")
library(caret)
## Loading required package: lattice
## Loading required package: ggplot2
library(randomForest) #Random forest for classification and regression

## randomForest 4.6-10
## Type rfNews() to see new features/changes/bug fixes.


library(rpart) # Regressive Partitioning and Regression trees
library(rpart.plot) # Decision Tree plot

# setting the overall seed for reproduceability
set.seed(1234)

```

###Loading data sets and preliminary cleaning

First we want to load the data sets into R and make sure that missing values are coded correctly.
Irrelevant variables will be deleted.
Results will be hidden from the report for clarity and space considerations.

```{r}

# After saving both data sets into my working directory
# Some missing values are coded as string "#DIV/0!" or "" or "NA" - these will be changed to NA.
# We notice that both data sets contain columns with all missing values - these will be deleted.  

# Loading the training data set into my R session replacing all missing with "NA"
trainingset <- read.csv("C:/Users/A6TTXZZ/Desktop/R Scripts/Coursera Assignments/Practical Machine Learning/pml-training.csv", na.strings=c("NA","#DIV/0!", ""))

# Loading the testing data set 
testingset <- read.csv('C:/Users/A6TTXZZ/Desktop/R Scripts/Coursera Assignments/Practical Machine Learning/pml-testing.csv', na.strings=c("NA","#DIV/0!", ""))


# Check dimensions for number of variables and number of observations
dim(trainingset)
dim(testingset)

# Delete columns with all missing values
trainingset<-trainingset[,colSums(is.na(trainingset)) == 0]
testingset <-testingset[,colSums(is.na(testingset)) == 0]

# Some variables are irrelevant to our current project: user_name, raw_timestamp_part_1, raw_timestamp_part_,2 cvtd_timestamp, new_window, and  num_window (columns 1 to 7). We can delete these variables.
trainingset   <-trainingset[,-c(1:7)]
testingset <-testingset[,-c(1:7)]

# and have a look at our new datasets:
dim(trainingset)
dim(testingset)
head(trainingset)
head(testingset)

```

###Partitioning the training data set to allow cross-validation

The training data set contains 53 variables and 19622 obs.
The testing data set contains 53 variables and 20 obs.
In order to perform cross-validation, the training data set is partionned into 2 sets: subTraining (75%) and subTest (25%).
This will be performed using random subsampling without replacement.

```{r}
subsamples <- createDataPartition(y=trainingset$classe, p=0.75, list=FALSE)
subTraining <- trainingset[subsamples, ] 
subTesting <- trainingset[-subsamples, ]
dim(subTraining)
dim(subTesting)
head(subTraining)
head(subTesting)


```

###A look at the Data

The variable "classe" contains 5 levels: A, B, C, D and E. A plot of the outcome variable will allow us to see the frequency of each levels in the subTraining data set and compare one another.

```{r}

plot(subTraining$classe, col="blue", main="Bar Plot of levels of the variable classe within the subTraining data set", xlab="classe levels", ylab="Frequency")

```

From the graph above, we can see that each level frequency is within the same order of magnitude of each other. Level A is the most frequent with more than 4000 occurrences while level D is the least frequent with about 2500 occurrences.

```{r}

model1 <- rpart(classe ~ ., data=subTraining, method="class")

# Predicting:
prediction1 <- predict(model1, subTesting, type = "class")

# Plot of the Decision Tree
rpart.plot(model1, main="Classification Tree", extra=102, under=TRUE, faclen=0)

# Test results on our subTesting data set:
confusionMatrix(prediction1, subTesting$classe)


model2 <- randomForest(classe ~. , data=subTraining, method="class")

# Predicting:
prediction2 <- predict(model2, subTesting, type = "class")

# Test results on subTesting data set:
confusionMatrix(prediction2, subTesting$classe)


```

As expected, Random Forest algorithm performed better than Decision Trees.
Accuracy for Random Forest model was 0.995 (95% CI: (0.993, 0.997)) compared to 0.739 (95% CI: (0.727, 0.752)) for Decision Tree model. The random Forest model is choosen. The accuracy of the model is 0.995. The expected out-of-sample error is estimated at 0.005, or 0.5%. The expected out-of-sample error is calculated as 1 - accuracy for predictions made against the cross-validation set. Our Test data set comprises 20 cases. With an accuracy above 99% on our cross-validation data, we can expect that very few, or none, of the test samples will be missclassified.

```{r}

# predict outcome levels on the original Testing data set using Random Forest algorithm
predictfinal <- predict(model2, testingset, type="class")
predictfinal


# Write files for submission
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(predictfinal)



```
```

